---
output: github_document
---

```{r global, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.path = "README_files/figure-gfm/",
                      dpi = 110, fig.width = 6, fig.height = 3.4)
```

# IBM Employee Attrition — The Cumulative Strain Hypothesis

**Research question:** How does the accumulation of work–life strain factors
(overtime, business travel, and marital status) affect employee attrition, and
can compensation mitigate it?

**Authors:** Tzlil Hayne, Oded Shmuely, Amit Shaimen, Topaz Sarid (SISE2601)

**Data:** the public IBM HR Analytics Employee Attrition dataset
([Kaggle](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset)).
To reproduce, place `WA_Fn-UseC_-HR-Employee-Attrition.csv` in a `data/` folder.

# Environment Setup

## Load Libraries

```{r libraries}
library(tidyverse)
library(broom)

navy <- "#1F4E5F"; accent <- "#C0622E"; grayc <- "grey60"
theme_set(theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank()))
```

## Read Data

```{r read-data}
ibm <- read_csv("data/WA_Fn-UseC_-HR-Employee-Attrition.csv", show_col_types = FALSE)
dim(ibm)
```

## Prepare Variables

```{r prepare}
# Drop constant / identifier columns, code the target and the strain factors
ibm <- ibm %>%
  select(!c(EmployeeCount, StandardHours, Over18, EmployeeNumber)) %>%
  mutate(
    Attr           = as.integer(Attrition == "Yes"),
    MaritalStatus  = factor(MaritalStatus, levels = c("Single", "Married", "Divorced")),
    BusinessTravel = factor(BusinessTravel,
                            levels = c("Non-Travel", "Travel_Rarely", "Travel_Frequently")),
    s_overtime = as.integer(OverTime == "Yes"),
    s_travel   = as.integer(BusinessTravel != "Non-Travel"),
    s_single   = as.integer(MaritalStatus == "Single"),
    StrainLoad = s_overtime + s_travel + s_single)            # cumulative strain score 0-3
ibm$IncomeBand <- factor(ntile(ibm$MonthlyIncome, 3),
                         labels = c("Low pay", "Medium pay", "High pay"))

base_rate  <- mean(ibm$Attr)                                  # overall attrition rate
strain_tab <- ibm %>% group_by(StrainLoad) %>%                # rates by strain score
  summarise(n = n(), left = sum(Attr), rate = mean(Attr), .groups = "drop")
base_rate
```

# Data Analysis

#part 2
#StrainLoad fig
```{r}
library(tidyverse)

strain_tab <- ibm %>% group_by(StrainLoad) %>%
  summarise(n = n(), left = sum(Attr), rate = mean(Attr), .groups = "drop")

ggplot(strain_tab, aes(factor(StrainLoad), rate)) +
  geom_col(fill = "#A7C8D6", width = 0.45) +
  geom_text(aes(label = paste0(scales::percent(rate, accuracy = 0.1), "\n(n=", n, ")")),
            vjust = -0.3, size = 3, lineheight = 0.9) +
  geom_hline(yintercept = mean(ibm$Attr), linetype = "dashed", colour = "grey60") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.6)) +
  labs(x = "Number of strain markers (0-3)", y = "Attrition rate")
```


```{r}
m_load <- glm(Attr ~ StrainLoad, binomial, ibm)
b  <- coef(m_load)["StrainLoad"]
se <- summary(m_load)$coefficients["StrainLoad", "Std. Error"]
ci <- exp(c(b - 1.96*se, b + 1.96*se))                       

trend <- prop.trend.test(strain_tab$left, strain_tab$n)        
m_sep <- glm(Attr ~ s_overtime + s_travel + s_single, binomial, ibm)
lrt   <- anova(m_load, m_sep, test = "LRT")

tibble(
  Statistic = c("OR per additional strain marker",
                "Cochran-Armitage trend test",
                "LRT: separate weights vs. simple count"),
  Result = c(sprintf("%.2f (95%% CI %.2f-%.2f), p < 0.001", exp(b), ci[1], ci[2]),
             sprintf("chi-square = %.1f, p < 0.001", trend$statistic),
             sprintf("p = %.2f (no improvement)", lrt$`Pr(>Chi)`[2]))
) %>% knitr::kable()
```

#part 4
# correlation income ODS-Ratios
```{r income-seniority-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(tidyverse)
library(knitr)

data_path <- "data/WA_Fn-UseC_-HR-Employee-Attrition.csv"

ibm <- read_csv(data_path, show_col_types = FALSE) %>%
  select(-EmployeeCount, -StandardHours, -Over18, -EmployeeNumber) %>%
  mutate(
    Attr = if_else(Attrition == "Yes", 1, 0)
  )

m_income <- glm(
  Attr ~ scale(MonthlyIncome),
  data = ibm,
  family = binomial
)

m_income_age <- glm(
  Attr ~ scale(MonthlyIncome) + Age,
  data = ibm,
  family = binomial
)

m_income_seniority <- glm(
  Attr ~ scale(MonthlyIncome) + Age + JobLevel + TotalWorkingYears,
  data = ibm,
  family = binomial
)

extract_income_effect <- function(model, model_name) {
  
  coef_table <- summary(model)$coefficients
  
  estimate <- coef_table["scale(MonthlyIncome)", "Estimate"]
  se <- coef_table["scale(MonthlyIncome)", "Std. Error"]
  p_value <- coef_table["scale(MonthlyIncome)", "Pr(>|z|)"]
  
  or <- exp(estimate)
  ci_low <- exp(estimate - 1.96 * se)
  ci_high <- exp(estimate + 1.96 * se)
  
  tibble(
    Model = model_name,
    `Income OR` = sprintf("%.2f", or),
    `95% CI` = paste0(sprintf("%.2f", ci_low), "–", sprintf("%.2f", ci_high)),
    `p-value` = if_else(
      p_value < 0.001,
      "<0.001",
      sprintf("%.3f", p_value)
    )
  )
}

income_seniority_table <- bind_rows(
  extract_income_effect(m_income, "Income only"),
  extract_income_effect(m_income_age, "Income + Age"),
  extract_income_effect(m_income_seniority, "Income + seniority controls")
)

cat('
<style>
.income-table-title {
  text-align: center;
  color: black;
  font-weight: bold;
  font-size: 11px;
  margin-top: 3px;
  margin-bottom: 4px;
}

.income-table {
  margin-left: auto;
  margin-right: auto;
  border-collapse: collapse;
  width: 46%;
  max-width: 430px;
  font-size: 11px;
  color: black;
}

.income-table th {
  color: black;
  font-weight: bold;
  text-align: center;
  padding: 5px 8px;
  border-bottom: 2px solid #D9D9D9;
  background-color: white;
}

.income-table td {
  color: black;
  text-align: center;
  padding: 5px 8px;
  border-bottom: 1px solid #E0E0E0;
  background-color: white;
}

.income-table th:first-child,
.income-table td:first-child {
  text-align: left;
  padding-left: 6px;
}

.income-table tr:last-child td {
  border-bottom: none;
}
</style>

<div class="income-table-title">
Income OR after seniority controls
</div>
')

income_seniority_table_compact <- income_seniority_table %>%
  mutate(
    Model = recode(
      Model,
      "Income only" = "Income only",
      "Income + Age" = "Income + Age",
      "Income + seniority controls" = "Income + seniority controls"
    )
  )

kable(
  income_seniority_table_compact,
  format = "html",
  escape = FALSE,
  align = c("l", "c", "c", "c"),
  table.attr = 'class="income-table"'
)
```


