---
output: github_document
---

```{r global, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.path = "README_files/figure-gfm/",
                      dpi = 110, fig.width = 6, fig.height = 3.4)
```

# IBM Employee Attrition — The Cumulative Strain Hypothesis

**Research question:** How does the accumulation of work–life strain factors
(overtime, business travel, and marital status) affect employee attrition, and
can compensation mitigate it?

**Authors:** Tzlil Hayne, Oded Shmuely, Amit Shaimen, Topaz Sarid (SISE2601)

**Data:** the public IBM HR Analytics Employee Attrition dataset
([Kaggle](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset)).
To reproduce, place `WA_Fn-UseC_-HR-Employee-Attrition.csv` in a `data/` folder.

# Environment Setup

## Load Libraries

```{r libraries}
library(tidyverse)
library(broom)

navy <- "#1F4E5F"; accent <- "#C0622E"; grayc <- "grey60"
theme_set(theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank()))
```

## Read Data

```{r read-data}
ibm <- read_csv("data/WA_Fn-UseC_-HR-Employee-Attrition.csv", show_col_types = FALSE)
dim(ibm)
```

## Prepare Variables

```{r prepare}
# Drop constant / identifier columns, code the target and the strain factors
ibm <- ibm %>%
  select(!c(EmployeeCount, StandardHours, Over18, EmployeeNumber)) %>%
  mutate(
    Attr           = as.integer(Attrition == "Yes"),
    MaritalStatus  = factor(MaritalStatus, levels = c("Single", "Married", "Divorced")),
    BusinessTravel = factor(BusinessTravel,
                            levels = c("Non-Travel", "Travel_Rarely", "Travel_Frequently")),
    s_overtime = as.integer(OverTime == "Yes"),
    s_travel   = as.integer(BusinessTravel != "Non-Travel"),
    s_single   = as.integer(MaritalStatus == "Single"),
    StrainLoad = s_overtime + s_travel + s_single)            # cumulative strain score 0-3
ibm$IncomeBand <- factor(ntile(ibm$MonthlyIncome, 3),
                         labels = c("Low pay", "Medium pay", "High pay"))

base_rate  <- mean(ibm$Attr)                                  # overall attrition rate
strain_tab <- ibm %>% group_by(StrainLoad) %>%                # rates by strain score
  summarise(n = n(), left = sum(Attr), rate = mean(Attr), .groups = "drop")
base_rate
```

# Data Analysis

